---
title: "Using LLMs in R with ellmer and NRP"
format:
  html:
    toc: true
    code-fold: false
---

This tutorial introduces [ellmer](https://ellmer.tidyverse.org/), the tidyverse R package for working with large language models, using the [NRP managed LLM service](https://nrp.ai/documentation/userdocs/ai/llm-managed/) as our backend. The NRP endpoint is OpenAI-compatible, provides access to several frontier open-weight models, and is free to use for researchers.

## Setup

You'll need an API key for the NRP endpoint. Store it as `ESPM_288_KEY` in your environment (e.g. in `~/.bashrc`):

```bash
export ESPM_288_KEY=your_key_here
```

Install ellmer if needed:

```r
install.packages("ellmer")
```

```{r}
#| message: false
library(ellmer)

api_key <- Sys.getenv("ESPM_288_KEY")
if (nchar(api_key) == 0) stop("ESPM_288_KEY not set")
```

## Connecting to the NRP endpoint

The NRP LLM service runs at `https://ellm.nrp-nautilus.io/v1` and exposes an OpenAI-compatible API, so we use `chat_openai_compatible()`. We'll wrap it in a helper that keeps the connection details in one place:

```{r}
nrp_chat <- function(model = "qwen3-small", ...) {
  chat_openai_compatible(
    base_url = "https://ellm.nrp-nautilus.io/v1",
    model = model,
    credentials = function() api_key,
    ...
  )
}
```

Available models include `qwen3`, `qwen3-small`, `gpt-oss`, `minimax-m2`, and `kimi`. We'll use `qwen3-small` (Qwen 3.5 27B) throughout — it's fast and capable.

## Basic chat

Create a chat object and call `$chat()`:

```{r}
chat <- nrp_chat(
  system_prompt = "You are a helpful assistant. Be concise.",
  echo = "output"
)

chat$chat("What is R, in one sentence?")
```

The chat object is stateful — each call to `$chat()` appends to the conversation history, so follow-up questions have context:

```{r}
chat$chat("What packages does it use for data visualization?")
```

## Tool calling

Tool calling lets the model invoke R functions to answer questions it couldn't otherwise answer — live data, computations, database lookups, etc. You define a function, register it with `register_tool()`, and the model decides when to call it.

Here we give the model a function that looks up approximate species counts by ecosystem:

```{r}
get_species_count <- function(ecosystem) {
  counts <- c(rainforest = 40000, coral_reef = 4000, temperate_forest = 6000)
  n <- counts[tolower(gsub(" ", "_", ecosystem))]
  if (is.na(n)) "unknown" else paste(n, "species")
}

chat2 <- nrp_chat(echo = "output")
chat2$register_tool(tool(
  get_species_count,
  description = "Returns approximate species count for a named ecosystem type.",
  arguments = list(
    ecosystem = type_string("The ecosystem type, e.g. 'rainforest', 'coral reef'.")
  )
))

chat2$chat("How many species live in a coral reef compared to a rainforest?")
```

The model called `get_species_count()` twice — once per ecosystem — then synthesized the results. The `◯` and `●` markers show the tool call and its return value.

## Structured data extraction

`$chat_structured()` guarantees the response matches a schema you define using `type_*` functions. This is useful for parsing text into data structures you can work with programmatically.

::: {.callout-important}
Most open-weight models need an explicit system prompt for extraction tasks. Without it, models respond *conversationally* ("Hi, I'm an AI assistant!") instead of extracting from the text — producing hallucinated names, function-style identifiers, or their own model name in the output fields. Adding a short extraction system prompt fixes this reliably across all NRP models.
:::

```{r}
extraction_prompt <- "Extract the requested fields from the user's text. Do not respond conversationally."
```

### Extracting a single record

```{r}
chat3 <- nrp_chat(system_prompt = extraction_prompt)

result <- chat3$chat_structured(
  "The golden poison dart frog (Phyllobates terribilis) lives in Colombian
   rainforests, grows to about 5.5 cm, and is listed as endangered.",
  type = type_object(
    common_name     = type_string("Common name of the species"),
    scientific_name = type_string("Scientific (Latin) name"),
    habitat         = type_string("Primary habitat"),
    length_cm       = type_number("Body length in centimetres"),
    status          = type_enum(
      "IUCN conservation status",
      values = c("least concern", "vulnerable", "endangered", "critically endangered")
    )
  )
)

str(result)
```

### Extracting a table

For tabular output, define the schema as an array of objects — ellmer will return a data frame:

```{r}
chat4 <- nrp_chat(system_prompt = extraction_prompt)

species_type <- type_array(
  "A list of species",
  items = type_object(
    common_name     = type_string("Common name"),
    scientific_name = type_string("Scientific name"),
    habitat         = type_string("Primary habitat"),
    endangered      = type_boolean("Whether the species is endangered")
  )
)

df <- chat4$chat_structured(
  "List three well-known endangered amphibian species with their habitats.",
  type = species_type
)

df
```

### Extracting from many prompts in parallel

`parallel_chat_structured()` runs the same schema extraction across a list of prompts concurrently, returning a tidy data frame where each row corresponds to one input:

```{r}
prompts <- list(
  "I go by Alex. 42 years on this planet and counting.",
  "Pleased to meet you! I'm Jamal, age 27.",
  "They call me Li Wei. Nineteen years young.",
  "Fatima here. Just celebrated my 35th birthday last week.",
  "The name's Robert - 51 years old and proud of it.",
  "Kwame here - just hit the big 5-0 this year."
)

type_person <- type_object(
  name = type_string("The person's name"),
  age  = type_number("The person's age in years")
)

chat5 <- nrp_chat(system_prompt = extraction_prompt)
parallel_chat_structured(chat5, prompts, type = type_person)
```

::: {.callout-note}
The extraction system prompt is essential here. Without it, models answer *as* an AI assistant rather than extracting from each prompt — substituting their own name and a default age, or inventing function-style identifiers like `greet_user`. This behaviour was confirmed across minimax-m2, gpt-oss, kimi, and qwen3 on the NRP endpoint; `qwen3-small` is the only model that extracts correctly without the prompt.
:::

## Thinking mode

Several NRP models — including `qwen3-small` — have extended reasoning ("thinking") enabled by default. The model works through the problem internally before responding, which improves accuracy on complex tasks but adds latency.

You can disable thinking by passing `chat_template_kwargs` via `api_args`. This maps directly to the `extra_body` parameter in the Python OpenAI client:

```{python}
#| eval: false
# Python equivalent
client.chat.completions.create(
  model="qwen3-small",
  extra_body={"chat_template_kwargs": {"enable_thinking": False}},
  ...
)
```

```{r}
#| label: thinking-on
prompt <- "How many times does the letter r appear in 'strawberry'?"

cat("--- Thinking ON (default) ---\n")
chat_think <- nrp_chat(echo = "output")
chat_think$chat(prompt)
```

```{r}
#| label: thinking-off
cat("--- Thinking OFF ---\n")
chat_no_think <- nrp_chat(
  echo = "output",
  api_args = list(chat_template_kwargs = list(enable_thinking = FALSE))
)
chat_no_think$chat(prompt)
```

::: {.callout-note}
Thinking tokens are returned in a separate `reasoning_content` field in the raw API response. ellmer does not currently display them, but their effect is visible in response latency and (often) answer quality on harder problems.
:::

The `kimi` model uses a slightly different parameter name:

```r
# For kimi model
nrp_chat(
  model = "kimi",
  api_args = list(chat_template_kwargs = list(thinking = FALSE))
)
```

## Vision (image input)

Several NRP models including `qwen3-small` are multimodal — they accept images alongside text. ellmer provides three helper functions for supplying image content:

- `content_image_url()` — a remote image URL
- `content_image_file()` — a local file path
- `content_image_plot()` — the current R plot device

Pass them as extra arguments to `$chat()`:

### Image from a file

Use `content_image_file()` for local images — ellmer base64-encodes the file and sends it inline. Use `content_image_url()` when the model server can reach the URL directly.

```{r}
# ellmer ships a sample PNG we can use for demonstration
sample_img <- system.file("httr2.png", package = "ellmer")

chat_vis <- nrp_chat(echo = "output")
chat_vis$chat(
  "What does this image show?",
  content_image_file(sample_img, resize = "none")
)
```

### Image from a plot

`content_image_plot()` captures whatever is currently on the R graphics device, making it easy to ask the model to interpret your data visualisations:

```{r}
# Create a plot, then ask the model to interpret it
plot(
  x = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020),
  y = c(2.5,  3.0,  3.7,  4.4,  5.3,  6.1,  6.9,  7.8),
  type = "b", pch = 19,
  xlab = "Year", ylab = "Population (billions)",
  main = "World Population Growth"
)

chat_plot <- nrp_chat(echo = "output")
chat_plot$chat(
  "Describe the trend shown in this plot and comment on what may have driven it.",
  content_image_plot()
)
```
