---
title: "NDVI and Redlining — Los Angeles"
subtitle: "Cloud-native zonal statistics via Planetary Computer, gdalcubes, and mapgl"
---

## Setup

```{r}
#| message: false
library(rstac)
library(gdalcubes)
library(duckdbfs)
library(sf)
library(dplyr)
library(mapgl)

gdalcubes_options(parallel = TRUE)
```

## 1. Load Los Angeles HOLC polygons

```{r}
holc <- open_dataset(
    "https://data.source.coop/cboettig/mappinginequality/mappinginequality.parquet"
) |>
    filter(city == "Los Angeles") |>
    select(-geom_bbox) |>
    rename(holc_fid = fid) |>
    to_sf(crs = 4326)

bbox <- st_bbox(holc)
bbox
```

## 2. Query Sentinel-2 via Planetary Computer STAC

Planetary Computer hosts Sentinel-2 L2A; items must be signed before use.

```{r}
items <- stac("https://planetarycomputer.microsoft.com/api/stac/v1/") |>
    stac_search(
        collections = "sentinel-2-l2a",
        bbox = as.numeric(bbox),
        datetime = "2025-06-01/2025-08-31",
        limit = 200
    ) |>
    ext_query("eo:cloud_cover" < 20) |>
    post_request() |>
    items_fetch() |>
    items_sign(sign_planetary_computer())

cat(length(items$features), "scenes found\n")
```

## 3. Build image collection

Planetary Computer Sentinel-2 L2A uses uppercase band names (`B04`, `B08`, `SCL`).

```{r}
col <- stac_image_collection(
    items$features,
    asset_names = c("B04", "B08", "SCL")
)
```

## 4. Define cube view and compute NDVI

```{r}
v <- cube_view(
    srs = "EPSG:4326",
    extent = list(
        t0 = "2025-06-01",
        t1 = "2025-08-31",
        left = bbox["xmin"],
        right = bbox["xmax"],
        bottom = bbox["ymin"],
        top = bbox["ymax"]
    ),
    dx = 0.0020,
    dy = 0.0020,
    dt = "P1M",
    aggregation = "median",
    resampling = "bilinear"
)
```


```{r}

# SCL values: 3 = cloud shadow, 8 = cloud medium, 9 = cloud high
S2.mask <- image_mask("SCL", values = c(3, 8, 9))

ndvi <- raster_cube(col, v, mask = S2.mask) |>
    select_bands(c("B04", "B08")) |>
    apply_pixel("(B08 - B04) / (B08 + B04)", "NDVI") |>
    reduce_time("mean(NDVI)")
```



```{r}
ndvi |>
    write_tif(
        dir = ".",
        prefix = "ndvi_la"
    )

```

::: {.callout-note}
`write_tif()` saves a GeoTIFF (e.g. `ndvi_la2025-06-01.tif`) to the current directory.
The file is ~2 MB and well within GitHub's 100 MB limit.
After rendering, commit and push the `.tif` alongside the notebook so the raster map below can load it via the GitHub Pages URL.
:::


```{r}
center <- c(mean(bbox[c("xmin", "xmax")]), mean(bbox[c("ymin", "ymax")]))

cog_url <- "https://espm-288.github.io/spatial-cboettig/ndvi_la2025-06-01.tif"
tilejson_url <- paste0(
    "https://titiler.nrp-nautilus.io/cog/WebMercatorQuad/tilejson.json",
    "?url=",
    URLencode(cog_url, reserved = TRUE),
    "&bidx=1&rescale=0,0.3&colormap_name=rdylgn"
)

maplibre(style = carto_style("positron"), center = center, zoom = 9) |>
    add_raster_source(
        id = "ndvi-raster",
        url = tilejson_url
    ) |>
    add_raster_layer(
        id = "ndvi-layer",
        source = "ndvi-raster",
        raster_opacity = 0.8
    ) |>
    add_legend(
        "Mean NDVI",
        values = c(0, 0.3),
        colors = c("#d73027", "#1a9850")
    )
```

## 5. Zonal statistics per HOLC parcel

```{r}
zonal <- ndvi |> extract_geom(holc, FUN = mean)
```

```{r}
holc$NDVI <- zonal$NDVI_mean

holc |>
    st_drop_geometry() |>
    group_by(grade) |>
    summarise(
        mean_NDVI = mean(NDVI, na.rm = TRUE),
        n = n()
    ) |>
    arrange(grade)
```

## 6. HOLC grade choropleth

```{r}
center <- c(mean(bbox[c("xmin", "xmax")]), mean(bbox[c("ymin", "ymax")]))

maplibre(style = carto_style("positron"), center = center, zoom = 10) |>
    add_fill_layer(
        id = "redlining",
        source = holc,
        fill_color = match_expr(
            column = "grade",
            values = list("A", "B", "C", "D"),
            stops = list("#76a865", "#7cb5bd", "#ffff00", "#d9838d")
        ),
        fill_opacity = 0.65,
        fill_outline_color = "white"
    ) |>
    add_legend(
        "HOLC Grade",
        values = c("A", "B", "C", "D"),
        colors = c("#76a865", "#7cb5bd", "#ffff00", "#d9838d"),
        type = "categorical"
    )
```

## 7. NDVI choropleth

```{r}
maplibre(style = carto_style("positron"), center = center, zoom = 10) |>
    add_fill_layer(
        id = "ndvi",
        source = holc,
        fill_color = interpolate(
            column = "NDVI",
            values = c(0, 0.14, 0.3),
            stops = c("#d73027", "#ffffbf", "#1a9850")
        ),
        fill_opacity = 0.8
    ) |>
    add_symbol_layer(
        id = "ndvi-labels",
        source = holc,
        text_field = get_column("grade"),
        text_size = 11,
        text_color = "white",
        text_halo_color = "rgba(0,0,0,0.6)",
        text_halo_width = 1.5
    ) |>
    add_legend(
        "Mean NDVI (summer 2023)",
        values = c(0, 0.3),
        colors = c("#d73027", "#1a9850")
    )
```

## 8. 3-D extrusion

```{r}
maplibre(
    style = carto_style("dark-matter"),
    center = center,
    zoom = 10,
    pitch = 50
) |>
    add_fill_extrusion_layer(
        id = "ndvi-3d",
        source = holc,
        fill_extrusion_color = interpolate(
            column = "NDVI",
            values = c(0, 0.6),
            stops = c("#d73027", "#1a9850")
        ),
        fill_extrusion_height = interpolate(
            column = "NDVI",
            values = c(0, 0.6),
            stops = c(0, 800)
        ),
        fill_extrusion_opacity = 0.85
    )
```
